---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


```{r message = FALSE, echo = FALSE}
library(dplyr)
```


## Overview

You can use the `stormwindmodel` package to model wind speeds at grid points in the United States based on "best tracks" hurricane tracking data. 

This package is currently in development on GitHub. You can install it using (you will need to install `devtools` if you do not already have it): 

```{r eval = FALSE}
devtools::install_github("geanders/stormwindmodel", build_vignettes = TRUE)
```

The package includes data on the tracks of Hurricane Floyd in 1999. You can load this example best tracks data using: 

```{r}
library(stormwindmodel)
data("floyd_tracks")
head(floyd_tracks)
```

This example data includes the following columns: 

- `storm_id`: Unique identifier for the tropical cyclone
- `date`: Date and time of the observation (in UTC)
- `latitude`, `longitude`: Location of the storm at that time
- `wind`: Maximum wind speed at that time (knots)

You must have your storm tracks in this format and with these columns names to input the tracks to the functions in `stormwindmodel`. If necessary, use `rename` from `dplyr` to rename columns and `convert_wind_speed` from `weathermetrics` to convert windspeed into knots.

The `stormwindmodel` package also includes a dataset with the location of the population mean center of each U.S. county (`county_points`). This dataset can be used as the grid point inputs if you want to model storm-related winds for counties. These counties are listed FIPS number. The population for [year] is also included. This dataset comes from the US Census [website].

```{r}
data(county_points)
head(county_points)
```


This package uses the wind model developed by Willoughby [reference] for modeling wind speed at each grid location. Full details on how this model is fit are provided in the "Details" vignetted of the `stormwindmodel` package.

## Basic example

The main function of this package is `get_grid_winds`. It inputs storm tracks for a tropical cyclone and a dataframe with grid point locations and outputs summaries of wind at each grid point from the storm. The wind measurements generated for each grid point are: 

- `vmax_gust`: Maximum 10-m 1-minute gust wind experienced at the grid point during the storm
- `vmax_sust`: Maximum 10-m 1-minute sustained wind experienced at the grid point during the storm
- `gust_dur`: Duration gust wind was at or above a specified speed (default is 20 m/s) in minutes
- `sust_dur`: Duration sustained wind was at or above a specified speed (default is 20 m/s) in minutes

To get modeled winds for Hurricane Floyd at U.S. county centers, you can run: 

```{r}
floyd_winds <- get_grid_winds(hurr_track = floyd_tracks,
                              grid_df = county_points)
floyd_winds %>%
  dplyr::select(gridid, vmax_gust, vmax_sust, gust_dur, sust_dur) %>%
  slice(1:6)
```

If you model winds for county centers, so the `gridid` is a county FIPS, the `stormwindmodel` package has a function called `map_wind` for mapping the estimated winds. By default, it maps the maximum sustained wind in each county during the storm in meters per second. 

```{r fig.width = 8}
map_wind(floyd_winds)
```

## Further functionality

There are a number of options when mapping wind speeds using `map_wind`. 

First, you can use the `map_tracks` function from `hurricaneexposure` package (also in development on GitHub) to add the storm track to the map. To do that, you save the wind map as an R object and input that object to `map_tracks` as the `plot_object`. This function can only be used for Atlantic basin storms between 1988 and 2015. 

```{r eval = FALSE}
install_github("geanders/hurricaneexposure")
```
```{r fig.width = 8}
library(hurricaneexposure)
floyd_map <- map_wind(floyd_winds)
map_tracks("Floyd-1999", plot_object = floyd_map)
```

You can also choose whether to map sustained or gust winds, as well as the unit to use for wind speed. 

```{r fig.width = 8}
map_wind(floyd_winds, value = "vmax_gust", wind_metric = "knots")
```

Finally, you can map a binary classification of counties with winds at or above a certain break point. For example, to map counties with sustained wind at or above 34 knots during the storm, you can run: 

```{r fig.width = 8}
map_wind(floyd_winds, value = "vmax_sust", wind_metric = "knots",
         break_point = 34)
```


## Tracks data

You can get an R version of this data for Atlantic basin storms from 1988 to 2015 through the `hurricaneexposuredata` package (also in development on GitHub):

```{r eval = FALSE}
devtools::install_github("geanders/hurricaneexposuredata")
```

```{r}
library(hurricaneexposuredata)
data("hurr_tracks")
hurr_tracks %>% 
  tidyr::separate(storm_id, c("storm", "year")) %>%
  dplyr::select(storm, year) %>%
  dplyr::distinct() %>%
  knitr::kable()
```

