---
title: "Modeling storm winds outside the United States"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling winds outside the United States}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  message = FALSE, 
  warning = FALSE
)
```

```{r setup}
library(stormwindmodel)
```

In the main vignettes for this package, we focused on modeling wind speeds
within counties in the Eastern half of the United States. However, you can also
use the wind model to model in other parts of the world, or to model at
different locations (rather than county centers) in the United States. This
vignette walks through some examples of how you can use the package to model
windspeeds any where.

## Creating a dataframe with grid points

You may have a dataframe with the latitude and longitude of the points you want
to model for a storm. For example, you may have a dataset with the location of
cities where you want to model winds, or you may have a shapefile with the
spatial information of another geography in your area of interest, which can be
used as a starting point to determine centroids, which you can then model.
However, if you do not have a starting dataset with locations to model, you can
still use this package. In this case, you'll want to start by creating a grid of
points at latitudes and longitudes that are evenly spaced across your area of
interest. This section of this vignette will walk through how you can do that
(if you already have your dataset of points to model to, you can skip this
section).

If you don't have a geographic dataset with locations to model winds, you can
create a gridded geographic dataset of the region of interest. The `sf` package 
has a function that can take an existing geographic R object and create a regularly-spaced
grid over the area of that object. You can use this to create your grid. 

First, to get the starting geographic R object, you can use the `rnaturalearth`
package, which provides spatial data for countries and subcountry geographies
around the world. You can use this to get the borders of countries worldwide,
then crop this object to the latitude and longitude ranges that you are
interested in. For example, the following code creates a cropped geographical R 
object for the region from 107 to 128 degrees longitude and 17.5 to 34 degrees latitude, 
focusing on an area affected by Typhoon Saomai in 2006:

```{r fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
library(rnaturalearth)
library(sf)
library(tidyverse)

all_countries <-  ne_countries(scale = "medium", 
                                  returnclass = "sf")
near_landfall <- all_countries %>% 
  st_make_valid() %>% 
  st_crop(y = st_bbox(c(xmin = 107, xmax = 128, ymin = 17.5, ymax = 34)))

ggplot() + 
  geom_sf(data = near_landfall) 
```

Next, you can use the `st_make_grid` function from the `sf` package to create
new objects with a regular grid over this area. You can use this function to
create two types of grids: centers---which will provide the point locations
(latitude and longitude) at the center of each grid---or polygons---which will
create geographic polygon shapes that cover the full grid. You will need the
point locations to input to the model, but you may also want to create a polygon
version to use to map the output later. Just make sure that, if you create both
types of objects, you use the same resolution for each (`cellsize`) and you
create a grid ID that aligns between the two datasets (which we do here with
`gridid = as.character(1:n())`) so you can connect the modeled winds back to the
polygon geographic data. Here is the code to create both types of objects (you can 
choose to have the grid shapes either be squares (`square = TRUE`) or hexagons 
(`square = FALSE`). We've used hexagons). The code creates a map showing both these 
layers, with transparency for the fill in the polygons (`alpha = 0.4`) so you can see 
all layers:

```{r fig.width = 6, fig.height = 5}

grid_near_landfall <- near_landfall %>% 
  st_make_grid(what = "polygons", square = FALSE, cellsize = 0.25) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))
points_near_landfall <- near_landfall %>% 
  st_make_grid(what = "centers", square = FALSE, cellsize = 0.25) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_sf(data = grid_near_landfall, color = "red", fill = NA, alpha = 0.4) + 
  geom_sf(data = points_near_landfall, color = "blue", fill = NA, alpha = 0.2, size = 0.5)
```

Next, you can use the point location geographic object to create a regular
dataframe that you can input to the wind model. This process includes extracting
the geometry information (latitude and longitude) into separate columns, which
you can do using the `st_coordinates` function from the `sf` package and then
"dropping" the geographic information from the object using `st_drop_geometry`
(this will change the object to a regular dataframe, rather than a spatial one).
It also involves using the land/sea mask from the package to determine which
points are over land and which are over water, which you can do using the
`check_land_function` from the `stormwindmodel` package. You can convert this to
a format that's ready to enter in the storm wind modeling function:

```{r}
points_to_model <- points_near_landfall %>% 
  mutate(glon = st_coordinates(.)[,1],
         glat = st_coordinates(.)[,2]) %>% 
  st_drop_geometry() %>% 
  mutate(gridid = gridid, 
         glandsea = mapply(check_over_land, tclat = .$glat, tclon = .$glon))

head(points_to_model)
```

You can re-join this dataframe with the spatial version and plot it to check
that the land/sea mask did a reasonable job of capturing grid points that are
over land versus over water:

```{r fig.width = 6, fig.height = 5}
points_near_landfall <- points_near_landfall %>%  
  left_join(points_to_model, by = "gridid")

ggplot() + 
  geom_sf(data = all_countries, fill = "tan") + 
  geom_sf(data = points_near_landfall, aes(color = glandsea), size = 0.8, alpha = 0.5) + 
  scale_color_manual(values = c("lightblue", "darkgreen"), guide = "none") + 
  coord_sf(xlim = c(107, 128), ylim = c(17.5, 34))
```

If you would like a better resolution of land / sea than is provided by this
landmask, you can use your own landmask dataset, with a high resolution focused 
on your area of interest, to create the `glandsea` column that you need in the grid
points dataset. Alternatively, if you are only modeling locaitons that are over 
land (e.g., only modeling at city centers), you could set this column to always
have the value `TRUE`. 

## Preparing storm tracks data for wind model function

The next step is to collect data on the central track and wind speeds of the
storm that you would like to model. This data is available for storms worldwide
from IBTrACS, so we are using that source as an example here. You will need to
reformat the data some to prepare it to serve as input to the wind modeling
functions in this package.

There is an example dataset in the `stormwindmodel` package that includes a
storm, Typhoon Saomai, that affected the area that we developed an example grid
for in the last section. The dataset is included in the format it would be in if
you read in directly from a CSV file from IBTrACS---the only changes are that
(1) only a few of the columns have been selected, rather than all the columns
that are available in the IBTrACS data, and (2) the column names have been
changed from all uppercase to all lowercase.

Before you can input this to the wind model code, you'll need to make a few
changes to the dataset. First, the date is in a date-time format, and you'll
need to change it to a character in the form of, for example, "200608040300" for
August 4, 2006 at 3 AM Greenwich Mean Time (UTM). You should name this column
"date". Next, you'll need to convert the columns with latitude and longitude to
numeric data types and change these column names from "lat" to "latitude" and
from "lon" to "longitude" (one way you can do this is using `transmute`, which
will only keep the new columns). Finally, you need to convert the column with
wind data (in US conventions of [wind averaging period and height]) to a numeric
data type and rename the column from "usa_wind" to "wind". You can do all these
steps with the following code:

```{r}
data("saomai_tracks")

saomai_to_model <- saomai_tracks %>% 
  transmute(date = format(iso_time, format = "%Y%m%d%H%M"), 
            latitude = as.numeric(lat), 
            longitude = as.numeric(lon),
            wind = as.numeric(usa_wind))
```

Now you have converted all the required data into the format that is needed to run the storm 
wind code. If you are interested in getting just the maximum wind speeds at each grid point (that
is, the maximum wind at that point over the full course of the storm), you can use the 
`get_grid_winds` function to do that. First, you'll run the function using the dataset of 
grid points that you created as the `grid_df` input and the track data for Typhoon Saomai as
the `hurr_track` input: 

```{r}
grid_winds <- get_grid_winds(hurr_track = saomai_to_model,
                             grid_df = points_to_model)

head(grid_winds)
```

The output will have a row for every grid point, where the grid point IDs match those from 
the grid point dataset you created (you can use this ID to link the data back to the spatial 
data object, for example, so you can plot the winds). For each grid point, it gives the maximum 
surface-level sustained windspeed over the course of the storm (`vmax_sust`), the maximum 
gust (`vmax_gust`---note that this is obtained as a simple multiplicative factor of `vmax_sust`), 
the duration of sustained and gust winds over 20 m/s (`sust_dur` and `gust_dur`, respectively; 
note that if you want to change the threshold from 20 m/s to another value, you can do so with 
the `sust_duration_cut` and `gust_duration_cut` parameters in `get_grid_winds`), and the date and
time (given based on the Greenwich Mean Time [UTM] time zone) when the wind was at its 
maximum at that grid point (typically, this will be around the time the storm was closest to 
the grid point, but there may be some cases where it's not the closest time point---for example, 
since the storm's wind often decays rapidly after landfall, a local wind may be highest while
the storm is still over water, even if the storm later passes closer to the location).

To plot these data, you can join them back with the spatial gridded dataframe, using the grid ID
to match between the two datasets. This gets the data back into an `sf` object type, which makes
it easy to map:

```{r}
modeled_near_landfall <- grid_near_landfall %>%  
  left_join(grid_winds, by = "gridid")

modeled_near_landfall %>% 
  head()
```

```{r fig.width = 6, fig.height = 5}
library(viridis)
ggplot() + 
  geom_sf(data = modeled_near_landfall, aes(color = vmax_sust, fill = vmax_sust)) + 
  geom_sf(data = all_countries, fill = NA, color = "white") + 
  scale_color_viridis(option = "B") + 
  scale_fill_viridis(option = "B") +   
  coord_sf(xlim = c(107, 128), ylim = c(17.5, 34))
```

You can zoom this map in by changing the ranges on the `coord_sf` layer, for example: 

```{r fig.width = 6, fig.height = 5}
ggplot() + 
  geom_sf(data = modeled_near_landfall, aes(color = vmax_sust, fill = vmax_sust)) + 
  geom_sf(data = all_countries, fill = NA, color = "white") + 
  scale_color_viridis(option = "B") + 
  scale_fill_viridis(option = "B") +   
  coord_sf(xlim = c(116, 123), ylim = c(25, 30))
```

You likely will also want to add the storm track. You have all the information you need to do 
that, since you have latitude and longitude of the storm track in the `saomai_to_model` object. 
However, to add it to the map, you'll need to convert it to an sf object class by specifying which
columns provide this information as well as giving an appropriate coordinate reference system 
number, so that it can be projected correctly. 

```{r fig.width = 6, fig.height = 5}
saomai_sf_points <- saomai_to_model %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(modeled_near_landfall)) 
saomai_sf_line <- saomai_sf_points %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

ggplot() + 
  geom_sf(data = modeled_near_landfall, aes(color = vmax_sust, fill = vmax_sust)) + 
  geom_sf(data = all_countries, fill = NA, color = "white") + 
  geom_sf(data = saomai_sf_line, color = "green") + 
  geom_sf(data = saomai_sf_points, color = "green") + 
  scale_color_viridis(option = "B") + 
  scale_fill_viridis(option = "B") +   
  coord_sf(xlim = c(107, 128), ylim = c(17.5, 34))

ggplot() + 
  geom_sf(data = modeled_near_landfall, aes(color = vmax_sust, fill = vmax_sust)) + 
  geom_sf(data = all_countries, fill = NA, color = "white") + 
  geom_sf(data = saomai_sf_line, color = "green") + 
  geom_sf(data = saomai_sf_points, color = "green") + 
  scale_color_viridis(option = "B") + 
  scale_fill_viridis(option = "B") +   
  coord_sf(xlim = c(116, 123), ylim = c(25, 30))
```

## Another example, from start to finish

Here is an example of the entire process for a second storm, Typhoon Mangkhut in 2018. Again, 
there is an example dataset that comes with the `stormwindmodel` package with this storm's 
tracks. For other storms, you would instead start form data downloaded from IBTrACS (for 
example, one of their CSV-format datasets). The gridded dataset is cropped in this case to 
focus on the area near the storm's landfall near Macau, Hong Kong, and Guangzhou. Since this
area is geographically more focused, in this example we use a smaller size for spacing between 
grid points.

```{r fig.width = 6, fig.height = 5}
data("mangkhut_tracks")

all_countries <-  ne_countries(scale = "medium", 
                                  returnclass = "sf")
near_landfall <- all_countries %>% 
  st_make_valid() %>% 
  st_crop(y = st_bbox(c(xmin = 110, xmax = 117, ymin = 19, ymax = 25)))

grid_near_landfall <- near_landfall %>% 
  st_make_grid(what = "polygons", square = FALSE, cellsize = 0.1) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))
points_near_landfall <- near_landfall %>% 
  st_make_grid(what = "centers", square = FALSE, cellsize = 0.1) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))

points_to_model <- points_near_landfall %>% 
  mutate(glon = sf::st_coordinates(.)[,1],
         glat = sf::st_coordinates(.)[,2]) %>% 
  st_drop_geometry() %>% 
  mutate(gridid = gridid, 
         glandsea = mapply(check_over_land, tclat = .$glat, tclon = .$glon))

points_near_landfall <- points_near_landfall %>%  
  left_join(points_to_model, by = "gridid")

ggplot() + 
  geom_sf(data = near_landfall, fill = "tan") + 
  geom_sf(data = points_near_landfall, aes(color = glandsea), size = 0.8, alpha = 0.5) + 
  scale_color_manual(values = c("lightblue", "darkgreen"), guide = "none") + 
  coord_sf(xlim = c(110, 117), ylim = c(19, 25))
```

In this case, you can see that there are a few areas where the land-sea mask has 
assigned grid points incorrectly near coastlines. This is likely because of the resolution 
of the land-sea mask available in the `stormwindmodel` package, which is constrained
to be a reasonable size to fit in a package. If you would like a finer resolution when 
determining whether a point is over land or sea, you can determine this for grid points 
using your own land-sea mask, at a higher resolution for the area you're modeling. Conversely,
if you are only modeling locations that are over land (for example, for a study of human 
impacts from a storm), then you can assign all grid points to "TRUE" for the `glandsea` 
column.

```{r fig.width = 6, fig.height = 5}
data("mangkhut_tracks")

mangkhut_to_model <- mangkhut_tracks %>% 
  transmute(date = format(iso_time, format = "%Y%m%d%H%M"), 
            latitude = as.numeric(lat), 
            longitude = as.numeric(lon),
            wind = as.numeric(usa_wind))

grid_winds <- get_grid_winds(hurr_track = mangkhut_to_model,
                             grid_df = points_to_model)

modeled_near_landfall <- grid_near_landfall %>%  
  left_join(grid_winds, by = "gridid")

mangkhut_sf_points <- mangkhut_to_model %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(modeled_near_landfall)) 
mangkhut_sf_line <- mangkhut_sf_points %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

ggplot() + 
  geom_sf(data = modeled_near_landfall, aes(color = vmax_sust, fill = vmax_sust)) + 
  geom_sf(data = all_countries, fill = NA, color = "white") + 
  geom_sf(data = mangkhut_sf_line, color = "green") + 
  geom_sf(data = mangkhut_sf_points, color = "green") + 
  scale_color_viridis(option = "B") + 
  scale_fill_viridis(option = "B") +   
  coord_sf(xlim = c(110, 117), ylim = c(19, 25))
```

