---
title: "Modeling windspeed and direction at a single location"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling windspeed and direction at a single location}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(stormwindmodel)
library(tidyverse)
```

This example focuses on how to model wind from a tropical cyclone at a 
single location, when the focus may be on a detailed description of the evolution
of wind speed and direction at that location over the course of the storm. 

As an example, we'll focus on a site in the Bermudas when it was hit by
Hurricane Dorian in 2019. We are modeling the winds at a location on Grand Abaco
Island (latitude: 26.45, longitude: -77.05). Since this location is very
low-lying, we've modeled it as being over water rather than over land (`glandsea
= FALSE`). We need to create a dataframe for this grid point, even though it 
will only be one row long (since there's only one location), because the 
storm modeling function is expecting a dataframe as the input for the location(s)
at which to model.

```{r}
location_to_model <- tibble(gridid = "Bermuda site, Grand Abaco Island", 
                            glat = 26.45, 
                            glon = -77.05, 
                            glandsea = FALSE)
location_to_model
```
Next, we need to get the track data for Hurricane Dorian. We have extracted that
tracking information from the IBTrACs dataset and saved it in a form in the
`stormwindmodel` package that is ready to use in the wind modeling function. 
You can load and use the data using the call: 

```{r}
data("dorian_tracks")

head(dorian_tracks)
```

At this point, we can map both the tracks of Dorian and the location where we'd
like to model the storm winds over the course of Dorian: 

```{r fig.width = 6, fig.height = 5, message = FALSE}
library(rnaturalearth)
library(sf)

# Create a map of land borders in the area
all_countries <-  ne_countries(scale = "medium", 
                                  returnclass = "sf")
near_landfall <- all_countries %>% 
  st_make_valid() %>% 
  st_crop(y = st_bbox(c(xmin = -82, xmax = -70, ymin = 20, ymax = 30)))

dorian_sf_points <- dorian_tracks %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(near_landfall)) 
dorian_sf_line <- dorian_sf_points %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

location_to_model_sf <- location_to_model %>% 
  st_as_sf(coords = c("glon", "glat"), crs = st_crs(near_landfall)) 

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_sf(data = dorian_sf_line, color = "red") + 
  geom_sf(data = dorian_sf_points, color = "darkred", size = 0.8) + 
  geom_sf(data = location_to_model_sf, color = "yellow") + 
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-82, -70), ylim = c(20, 30))
```

Here's another version, zoomed in to the study location and the color of 
points along the storm track used to show the storm's central maximum sustained 
wind at that time: 

```{r fig.width = 6, fig.height = 5}
library(viridis)
ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_sf(data = dorian_sf_line, color = "red") + 
  geom_sf(data = dorian_sf_points, aes(color = wind)) + 
  geom_sf(data = location_to_model_sf, color = "yellow", size = 3) + 
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-79, -76.5), ylim = c(25.8, 27.2)) + 
  scale_color_viridis(option = "A", direction = -1)
```

Next, we can run the wind modeling function: 

```{r fig.width = 8, fig.height = 4}
dorian_modeled <- calc_grid_winds(hurr_track = dorian_tracks, 
                                 grid_df = location_to_model)

str(dorian_modeled)

library(lubridate)
dorian_modeled %>% 
  pluck("vmax_sust") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(peak_wind = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  ggplot(aes(x = date, y = peak_wind)) + 
  geom_line() + 
  ylab("Modeled maximum sustained winds\nat study location (m/s)")
```

```{r fig.width = 6, fig.height = 5}
dorian_modeled %>% 
  pluck("vmax_sust") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(peak_wind = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  ggplot(aes(x = date, y = peak_wind)) + 
  geom_line() + 
  xlim(ymd_hms(c("2019-09-01 00:00:00", "2019-09-02 12:00:00"))) + 
  ylab("Modeled maximum sustained winds\nat study location (m/s)")
```

```{r fig.width = 6, fig.height = 5}
dorian_modeled %>% 
  pluck("surface_wind_direction") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(swd = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  ggplot(aes(x = date, y = swd)) + 
  geom_line() + 
  xlim(ymd_hms(c("2019-09-01 00:00:00", "2019-09-02 12:00:00"))) + 
  ylab("Surface wind direction (in polar conventions)")
```

```{r}
dorian_modeled %>% 
  pluck("surface_wind_direction") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(swd = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  filter(date %in% c(ymd_h("2019-09-01 14"), 
                     ymd_h("2019-09-01 15"),
                     ymd_h("2019-09-01 20"), 
                     ymd_h("2019-09-01 21")))
```

Next, we can map the modeled surface wind direction at regular intervals as
the storm passed it. For example, we could plot this every hour as the storm 
passed over the location.

```{r fig.height = 7, fig.width = 8}
modeled_wind_direction <- dorian_modeled %>% 
  pluck("surface_wind_direction") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(swd = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  filter(date %in% c(ymd_h("2019-09-01 14"),
                     ymd_h("2019-09-01 15"),
                     ymd_h("2019-09-01 16"),
                     ymd_h("2019-09-01 17"),
                     ymd_h("2019-09-01 18"),
                     ymd_h("2019-09-01 19"),
                     ymd_h("2019-09-01 20"),
                     ymd_h("2019-09-01 21"))) %>% 
  mutate(glat = location_to_model$glat, 
         glon = location_to_model$glon,
          mag = 1) 


modeled_wind_direction_sf <- dorian_modeled %>% 
  pluck("surface_wind_direction") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  rename(swd = `Bermuda site, Grand Abaco Island`) %>% 
  mutate(date = ymd_hms(date)) %>% 
  filter(date %in% c(ymd_h("2019-09-01 14"),
                     ymd_h("2019-09-01 15"),
                     ymd_h("2019-09-01 16"),
                     ymd_h("2019-09-01 17"),
                     ymd_h("2019-09-01 18"),
                     ymd_h("2019-09-01 19"),
                     ymd_h("2019-09-01 20"), 
                     ymd_h("2019-09-01 21"))) %>% 
  mutate(glat = location_to_model$glat, 
         glon = location_to_model$glon) %>% 
  st_as_sf(coords = c("glon", "glat"), crs = st_crs(near_landfall)) 


library("metR")
ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_sf(data = dorian_sf_line) + 
  geom_arrow(data = modeled_wind_direction %>% 
            filter(date == ymd_h("2019-09-01 20")), 
            aes(x = glon, y = glat, angle = swd, mag = mag), 
            color = "red", 
            direction = "ccw") + 
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-79, -76.5), ylim = c(25.8, 27.2)) + 
  scale_color_viridis(option = "A", direction = -1)
```


Make a grid with points throughout the area near landfall: 

```{r fig.height = 7, fig.width = 8}
near_lf_grid <- expand_grid(glon = seq(-79.0, -76.5, by = 0.1), 
                            glat = seq(25.5, 27.5, by = 0.1)) %>% 
  mutate(gridid = as.character(1:n()), 
         glandsea = FALSE)

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_point(data = near_lf_grid, aes(x = glon, y = glat), color = "yellow") + 
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-79, -76.5), ylim = c(25.5, 27.5))
```

```{r fig.height = 7, fig.width = 8}
dorian_modeled_2 <- calc_grid_winds(hurr_track = dorian_tracks, 
                                    grid_df = near_lf_grid)

modeled_wind_direction_2 <- dorian_modeled_2 %>% 
  pluck("surface_wind_direction") %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>%
  mutate(date = ymd_hms(date)) %>% 
  filter(date %in% c(ymd_h("2019-09-01 14"),
                     ymd_h("2019-09-01 15"),
                     ymd_h("2019-09-01 16"),
                     ymd_h("2019-09-01 17"),
                     ymd_h("2019-09-01 18"),
                     ymd_h("2019-09-01 19"),
                     ymd_h("2019-09-01 20"),
                     ymd_h("2019-09-01 21"))) %>% 
  pivot_longer(-date, names_to = "gridid", values_to = "swd") %>% 
  left_join(near_lf_grid, by = "gridid") %>% 
  mutate(mag = 0.2)

dorian_tracks_2 <- dorian_tracks %>% 
  mutate(date = ymd_hm(date)) 

dorian_tracks_3 <- dorian_tracks %>% 
  mutate(date = ymd_hm(date)) %>% 
  filter(date <= ymd_h("2019-09-01 15"))

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_path(data = dorian_tracks_2, aes(x = longitude, y = latitude), 
            color = "gray") + 
  geom_path(data = dorian_tracks_3, aes(x = longitude, y = latitude), 
            color = "red") + 
  geom_arrow(data = filter(modeled_wind_direction_2, 
                           date == ymd_h("2019-09-01 15")), 
            aes(x = glon, y = glat, angle = swd, mag = mag), 
            direction = "ccw",
            color = "yellow") + 
  geom_point(data = near_lf_grid, aes(x = glon, y = glat), size = 0.5) + 
  geom_arrow(data = modeled_wind_direction %>% 
            filter(date == ymd_h("2019-09-01 15")), 
            aes(x = glon, y = glat, angle = swd, mag = mag), 
            color = "blue", 
            direction = "ccw") +
  geom_point(data = modeled_wind_direction %>% 
            filter(date == ymd_h("2019-09-01 15")), 
            aes(x = glon, y = glat), size = 0.5) + 
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-79, -76.5), ylim = c(25.5, 27.5)) + 
  scale_mag(max_size = 0.5)

dorian_tracks_3 <- dorian_tracks %>% 
  mutate(date = ymd_hm(date)) %>% 
  filter(date <= ymd_h("2019-09-01 21"))

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_path(data = dorian_tracks_2, aes(x = longitude, y = latitude), 
            color = "gray") + 
  geom_path(data = dorian_tracks_3, aes(x = longitude, y = latitude), 
            color = "red") + 
  geom_arrow(data = filter(modeled_wind_direction_2, 
                           date == ymd_h("2019-09-01 21")), 
            aes(x = glon, y = glat, angle = swd, mag = mag), 
            direction = "ccw",
            color = "yellow") + 
  geom_point(data = near_lf_grid, aes(x = glon, y = glat), size = 0.5) + 
  geom_arrow(data = modeled_wind_direction %>% 
            filter(date == ymd_h("2019-09-01 21")), 
            aes(x = glon, y = glat, angle = swd, mag = mag), 
            color = "blue", 
            direction = "ccw") +
  theme(panel.background = element_rect(fill = "lightblue"))  + 
  coord_sf(xlim = c(-79, -76.5), ylim = c(25.5, 27.5)) + 
  scale_mag(max_size = 0.5)
```

