---
title: "Extracting snapshots of the wind field at different times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting snapshots of the wind field at different times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(stormwindmodel)
library(rnaturalearth)
library(sf)
library(tidyverse)
library(lubridate)
library(viridis)
```

In the other vignettes, we have focused on how to create summaries of how severe the peak
winds or wind gusts were at any point over the course of a storm for a certain location. 
In the course of modeling these storm-long values, the model estimates winds at a number
of time points over the course of the storm. This vignette describes how you can extract
and use those values to look at snapshots of the modeled wind field at specific times
during the storm. 

This example will use the Typhoon Mangkhut from 2018. The tracks for this storm are included
as an example dataset in this package. We have described how to set up these data for the 
model, as well as how to create a set of grid points to model to, in another vignette for
this package ("Modeling winds outside the United States"). We have repeated the code here, 
but please check the other vignette for more details on the process: 

```{r}
data("mangkhut_tracks")

all_countries <-  ne_countries(scale = "medium", 
                                  returnclass = "sf")
near_landfall <- all_countries %>% 
  st_make_valid() %>% 
  st_crop(y = st_bbox(c(xmin = 110, xmax = 117, ymin = 19, ymax = 25)))

grid_near_landfall <- near_landfall %>% 
  st_make_grid(what = "polygons", square = FALSE, cellsize = 0.1) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))
points_near_landfall <- near_landfall %>% 
  st_make_grid(what = "centers", square = FALSE, cellsize = 0.1) %>% 
  st_sf() %>% 
  mutate(gridid = as.character(1:n()))

points_to_model <- points_near_landfall %>% 
  mutate(glon = sf::st_coordinates(.)[,1],
         glat = sf::st_coordinates(.)[,2]) %>% 
  st_drop_geometry() %>% 
  mutate(gridid = gridid, 
         glandsea = mapply(check_over_land, tclat = .$glat, tclon = .$glon))

mangkhut_to_model <- mangkhut_tracks %>% 
  transmute(date = format(iso_time, format = "%Y%m%d%H%M"), 
            latitude = as.numeric(lat), 
            longitude = as.numeric(lon),
            wind = as.numeric(usa_wind))
```

Next, we can use these inputs to model the wind field during the storm. To generate output
that includes the values for each of the timepoints that were modeled, you can use the 
`...` function from the package: 

```{r}
mangkhut_winds <- calc_grid_winds(hurr_track = mangkhut_to_model, 
                                  grid_df = points_to_model)
```

The output from this function is a list, each of which contains a matrix. 
The modeled winds are included in the element of the list named `vmax_sust`. 
Here is a section pulled from the middle of this matrix: 

```{r}
mangkhut_winds$vmax_sust[900:910, 2500:2505]
```

You can see that each row gives data for a specific time, while each column gives
data for a specific grid point. 

You can use this matrix in a couple of ways. First, if you'd like to get a time series
of the modeled winds at a certain location over the course of a storm, you can do 
that. For example, we could create the time series of modeled winds in Hong Kong for
this storm. The latitude and longitude of Hong Kong are 22.32 degrees North and 114.17
degrees East. 

```{r}
points_to_model %>% 
  mutate(dist_hk_lat = abs(glat - 22.32), 
         dist_hk_lon = abs(glon - 114.17)) %>% 
  filter((dist_hk_lat + dist_hk_lon) == 
           min(dist_hk_lat + dist_hk_lon))
points_to_model %>% 
  filter(gridid == "2995")
```

```{r}
hk_winds <- mangkhut_winds$vmax_sust[ , "2995"] %>% 
  enframe(name = "date", value = "vmax_sust") %>% 
  mutate(date = ymd_hms(date))
hk_winds
```

```{r fig.width = 8, fig.height = 4}
ggplot(hk_winds, aes(x = date, y = vmax_sust)) + 
  geom_point()
```


```{r fig.width = 8, fig.height = 4}
mangkhut_timepoint <- mangkhut_winds$vmax_sust[rownames(mangkhut_winds$vmax_sust) %in%
                                                 c("2018-09-15 21:00:00", 
                                                   "2018-09-16 05:00:00", 
                                                   "2018-09-16 14:00:00"), ] %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "date") %>% 
  mutate(date = ymd_hms(date)) %>% 
  pivot_longer(-date, names_to = "gridid", values_to = "vmax_sust")

spatial_mangkhut_tp <- grid_near_landfall %>% 
  left_join(mangkhut_timepoint, by = "gridid")

ggplot() + 
  geom_sf(data = near_landfall) + 
  geom_sf(data = spatial_mangkhut_tp, aes(fill = vmax_sust), color = NA) + 
  scale_fill_viridis(direction = -1, option = "B") + 
  facet_wrap(~ format(date, "%Y-%m-%d %H:%M"))
```

